### 2025-06-28 lecturenotes_aima203/ch6_aima203.tex
ch 6

\begin{equation}
  \label{newtforwarddif}
  \begin{aligned}
    \frac{dy}{dx} =\frac{dy}{du} \frac{du}{dx} = \frac{1}{h} \left [ \Delta y_0 + \frac{2u -1}{2}\, \Delta^2\,  y_0 + \frac{3u^2-6u + 2}{6}\, \Delta^3 \, y_0 + ... \right ]
  \end{aligned}
\end{equation}
Differentiating ~\ref{newtforwarddif} with respect to $x$, we get,
\begin{equation}
  \label{newtforwarddif2}
  \begin{aligned}
    \frac{d^2y}{dx^2} = \frac{1}{h^2} \left [ \Delta^2 y_0 + \frac{6u -6}{6}\, \Delta^3 \,  y_0 + \frac{12u^2-36u + 22}{24}\, \Delta^4 \, y_0 + ... \right ]
  \end{aligned}
\end{equation}
These formulae are used for \textit{non-tabular} values of $x$. For tabular values of $x$, the formulae take simpler form. For $x=x_0$, we have $u=0$, and using this we can find the relations of $\displaystyle \left [ \frac{dy}{dx} \right ]  _{x=x_0} and \left [\frac{d^2y}{dx^2} \right ] _{x=x_0}$ which is in a simpler form.

\newpage
\subsection{Newton's backward difference formula}
In a similar way, different formulae can be derived by starting with other interpolation formulae. Thus, Newton's backward difference formula gives
\begin{equation}
  \label{newtbackwarddif}
  \left [ \frac{dy}{dx} \right ] _{x=x_n} = \frac{1}{h} \left [\nabla \, y_n + \frac{1}{2} \nabla ^2 \, y_n + \frac{1}{3} \nabla ^3 \, y_n + ... \right ]
\end{equation}
and

\begin{equation}
  \label{newtbackwarddif2}
  \left [ \frac{d^2y}{dx^2} \right ] _{x=x_n} = \frac{1}{h^2} \left [\nabla^2 \, y_n + \nabla ^3 \, y_n + \frac{11}{12} \nabla ^4 \, y_n + \frac{5}{6} \nabla ^5 \, y_n + ... \right ]
\end{equation}
\begin{remark}
   If the $x$-values of the data points are equally spaced then it is common to use Newton's interpolation.
\end{remark}
\subsection{Exercise}
Find the $dy/dx$ and $d^2y/dx^2$ at $x=1.2, \; x=1.6, \; x=2.2$ from the data: \\[1mm]
$(1, 2.7183), \; (1.2, 3.3201), \; (1.4, 4.0522), \; (1.6, 4.9530), \; (1.8, 6.0496), \; (2.0, 7.3891), \; (2.2, 9.0250)$.
\section{Numerical Integration}
The general problem of numerical integration may be stated as: Given a set of data points of a function $y=f(x)$, where $f(x)$ is not known explicitly, it is required compute the value of the definite integral $$I = \int_a^b \; y\,dx$$ As with the case of numerical differentiation, one replaces $f(x)$ by an interpolating polynomial $\phi(x)$ and obtains, on integration, an approximate value of the definite integral. Thus, different integration formulae can be obtained depending upon the type of the interpolation formula used. Approximating $y$ by Newton's forward difference formula, we obtained
$$ I = \int_{x_0}^{x_n} \; \left [ y_0 + p \Delta y_0 + \frac{p(p-1)}{2}\, \Delta^2 y_0 + \frac{p(p-1)(p-2)}{6}\, \Delta ^3 y_0 + ... \right ] \, dx$$
Since $x=x_0 +ph$, $dx = hdp$. When $x=x_0$, $p=0$ and $x=x_n$, $p=n$. Then,
$$ I = \int_{0}^{n} \; \left [ y_0 + p \Delta y_0 + \frac{p(p-1)}{2}\, \Delta^2 y_0 + \frac{p(p-1)(p-2)}{6}\, \Delta ^3 y_0 + ... \right ] \, h\,dp$$
Integrating we get,
\begin{equation}
  \label{numint}
  \begin{aligned}[b]
    I &= h \left [ py_0 + \frac{p^2}{2}\, \Delta y_0 + \left ( \frac{p^3}{3} - \frac{p^2}{2} \right ) \, \frac{\Delta^2 y_0}{2} +  \left ( \frac{p^4}{4} - p^3 + p^2 \right ) \frac{\Delta ^3 \, y_0}{6} + ... \right ]_0^n \\[1mm]
    &= nh \left [ y_0 + \frac{n}{2}\, \Delta y_0 + \left ( \frac{2n^2-3n}{12} \right ) \, \Delta^2\, y_0 +  \left ( \frac{n^3 - 4n^2 +4n}{24} \right ) \, \Delta ^3 \, y_0 + ... \right ]
  \end{aligned}
\end{equation}

@@

\subsection{Trapezoidal Rule}
Setting $n=1$ in the general formula ~\ref{numint}, all differences higher than the first will become zero and we obtained \\[1mm]
$\displaystyle \int_{x_0}^{x_1} \; y\,dx = h \left [ y_0 + \frac{1}{2} \Delta y_0 \right ] = h \left [ y_0 + \frac{1}{2}(y_1 -y_0) \right ] = \frac{h}{2}\, (y_0 + y_1).$ Similarly we can obtain \\[5mm] $\int_{x_1}^{x_2} \; y\,dx, ..... \int_{x_{n-1}}^n \; y\,dx $. Then we have,
\begin{equation}
  \label{trap}
  \begin{aligned}[b]
    \int_{x_0}^{x_n}\; y\,dx &= \int_{x_0}^{x_1} \; y\,dx + \int_{x_1}^{x_2} \; y\,dx + .... + \int_{x_{n-1}}^{x_n} \; y\,dx \\[1mm]
                             &= \frac{h}{2} \, (y_0 + y_1) + \frac{h}{2} \, (y_1 + y_2) + .... + \frac{h}{2} \, (y_{n-1} + y_n) \\[1mm]
                             &= \frac{h}{2} \, \{ y_0 + 2(y_1 + y_2 + ... + y_{n-1}) + y_n \}
  \end{aligned}
\end{equation}

@@

\subsection{Exercise}
\begin{enumerate}
\item Find, from the data, the area bounded by the curve and the $x$-axis: \\[1mm]
  $(7.47, 1.93), \; (7.48, 1.95),\; (7.49, 1.98),\; (7.50, 2.01),\; (7.51, 2.03),\; (7.52, 2.06)$

\item Find the volume of solid of revolution formed by rotating about the $x$-axis the area between the $x$-axis, the line $x=0$ and $x=1$, and a curve through the points: \\[1mm]
  $(0,1),\; (0.25,0.9896),\; (0.5,0.9589),\; (0.75,0.9089),\; (1,0.8415)$.4

\item Integrate $\displaystyle I = \int_0^{1.5}\; \frac{e^x +x}{sinx +1}\, dx$.

\item Evaluate: $\displaystyle I = \int_0^1 \; \frac{1}{1+x}\, dx$ using $h=0.125$.
\end{enumerate}

\section{Double Integration}
Formulae for the evaluation of double integral can be obtained by repeatedly applying the trapezoidal and Simpson's rules derived earlier.
\subsection{Trapezoidal Rule}
\begin{equation*}
  I = \int_{y_j}^{y_{j+1}}\; \int_{x_i}^{x_{i+1}} \; f(x,y)\;dx\,dy
\end{equation*}
where, $x_{i+1} = x_i + h$ and $y_{j+1} = y_j + k$.
By repeated application of trapezoidal rule to we get,
\begin{equation}
  \label{trapdouble}
  \begin{aligned}
    I &= \frac{h}{2} \,\int_{y_j}^{y_{j+1}}\; [f(x_i,y) + f(x_{i+1},y) ]\;dx\,dy \\[1mm]
      &= \frac{hk}{4} \, [f(x_i,y_j) + f(x_{i+1},y_j) + f(x_i,y_{j+1}) + f(x_{i+1},y_{j+1}) ]\;dx\,dy \\[1mm]
      &= \frac{hk}{4} \, [f_{i,j} + f_{i+1, j} + f_{i,j+1} + f_{i+1,j+1} ]\;dx\,dy
  \end{aligned}
\end{equation}

@@

\section{Romberg Integration}
This method can often be used to improve the approximate results obtained by the finite-difference methods. Its application to the numerical evaluation of definite integrals, for example in the use of trapezoidal rule, can be described, as follows.
\end{document}

>>>>

\begin{equation}
  \label{newtforwarddif}
  \begin{aligned}
    \frac{dy}{dx} =\frac{dy}{du} \frac{du}{dx} = \frac{1}{h} \left [ \Delta y_0 + \frac{2u -1}{2}\, \Delta^2\,  y_0 + \frac{3u^2-6u + 2}{6}\, \Delta^3 \, y_0 + ... \right ]
  \end{aligned}
\end{equation}
Differentiating ~\ref{newtforwarddif} with respect to $x$, we get,
\begin{equation}
  \label{newtforwarddif2}
  \begin{aligned}
    \frac{d^2y}{dx^2} = \frac{1}{h^2} \left [ \Delta^2 y_0 + \frac{6u -6}{6}\, \Delta^3 \,  y_0 + \frac{12u^2-36u + 22}{24}\, \Delta^4 \, y_0 + ... \right ]
  \end{aligned}
\end{equation}
These formulae are used for \textit{non-tabular} values of $x$. For tabular values of $x$, the formulae take simpler form. For $x=x_0$, we have $u=0$, and using this we can find the relations of $\displaystyle \left [ \frac{dy}{dx} \right ]  _{x=x_0} and \left [\frac{d^2y}{dx^2} \right ] _{x=x_0}$ which is in a simpler form.
\newpage
\begin{equation}
  \label{newtforw1x}
  \left [ \frac{dy}{dx} \right ] _{x=x_n} = \frac{1}{h} \left [\Delta \, y_n - \frac{1}{2} \Delta ^2 \, y_n + \frac{1}{3} \Delta ^3 \, y_n - ... \right ]
\end{equation}
and
\begin{equation}
  \label{newtforw2x}
  \left [ \frac{d^2y}{dx^2} \right ] _{x=x_n} = \frac{1}{h^2} \left [\Delta^2 \, y_n - \Delta ^3 \, y_n + \frac{11}{12} \Delta ^4 \, y_n - \frac{5}{6} \Delta ^5 \, y_n + ... \right ]
\end{equation}

\subsection{Newton's backward difference formula}
In a similar way, different formulae can be derived by starting with other interpolation formulae. Thus, Newton's backward difference formula gives
\begin{equation}
  \label{newtbackwarddif}
  \left [ \frac{dy}{dx} \right ] _{x=x_n} = \frac{1}{h} \left [\nabla \, y_n + \frac{1}{2} \nabla ^2 \, y_n + \frac{1}{3} \nabla ^3 \, y_n + ... \right ]
\end{equation}
and
\begin{equation}
  \label{newtbackwarddif2}
  \left [ \frac{d^2y}{dx^2} \right ] _{x=x_n} = \frac{1}{h^2} \left [\nabla^2 \, y_n + \nabla ^3 \, y_n + \frac{11}{12} \nabla ^4 \, y_n + \frac{5}{6} \nabla ^5 \, y_n + ... \right ]
\end{equation}
\begin{remark}
   If the $x$-values of the data points are equally spaced then it is common to use Newton's interpolation.
\end{remark}
\subsection{Exercise}
Find the $dy/dx$ and $d^2y/dx^2$ at $x=1.2, \; x=1.6, \; x=2.2$ from the data: \\[1mm]
$(1, 2.7183), \; (1.2, 3.3201), \; (1.4, 4.0522), \; (1.6, 4.9530), \; (1.8, 6.0496), \; (2.0, 7.3891), \; (2.2, 9.0250)$.
\section{Numerical Integration}
The general problem of numerical integration may be stated as: Given a set of data points of a function $y=f(x)$, where $f(x)$ is not known explicitly, it is required compute the value of the definite integral $$I = \int_a^b \; y\,dx$$ As with the case of numerical differentiation, one replaces $f(x)$ by an interpolating polynomial $\phi(x)$ and obtains, on integration, an approximate value of the definite integral. Thus, different integration formulae can be obtained depending upon the type of the interpolation formula used. Approximating $y$ by Newton's forward difference formula, we obtained
$$ I = \int_{x_0}^{x_n} \; \left [ y_0 + p \Delta y_0 + \frac{p(p-1)}{2}\, \Delta^2 y_0 + \frac{p(p-1)(p-2)}{6}\, \Delta ^3 y_0 + ... \right ] \, dx$$
Since $x=x_0 +ph$, $dx = hdp$. When $x=x_0$, $p=0$ and $x=x_n$, $p=n$. Then,
$$ I = \int_{0}^{n} \; \left [ y_0 + p \Delta y_0 + \frac{p(p-1)}{2}\, \Delta^2 y_0 + \frac{p(p-1)(p-2)}{6}\, \Delta ^3 y_0 + ... \right ] \, h\,dp$$
Integrating we get,
\begin{equation}
  \label{numint}
  \begin{aligned}[b]
    I &= h \left [ py_0 + \frac{p^2}{2}\, \Delta y_0 + \left ( \frac{p^3}{3} - \frac{p^2}{2} \right ) \, \frac{\Delta^2 y_0}{2} +  \left ( \frac{p^4}{4} - p^3 + p^2 \right ) \frac{\Delta ^3 \, y_0}{6} + ... \right ]_0^n \\[1mm]
    &= nh \left [ y_0 + \frac{n}{2}\, \Delta y_0 + \left ( \frac{2n^2-3n}{12} \right ) \, \Delta^2\, y_0 +  \left ( \frac{n^3 - 4n^2 +4n}{24} \right ) \, \Delta ^3 \, y_0 + ... \right ]
  \end{aligned}
\end{equation}

@@

\subsection{Trapezoidal Rule}
Setting $n=1$ in the general formula ~\ref{numint}, all differences higher than the first order will become zero and we obtained \\[1mm]
$\displaystyle \int_{x_0}^{x_1} \; y\,dx = h \left [ y_0 + \frac{1}{2} \Delta y_0 \right ] = h \left [ y_0 + \frac{1}{2}(y_1 -y_0) \right ] = \frac{h}{2}\, (y_0 + y_1).$ Similarly we can obtain \\[5mm] $\int_{x_1}^{x_2} \; y\,dx, ..... \int_{x_{n-1}}^n \; y\,dx $. Then we have,
\begin{equation}
  \label{trap}
  \begin{aligned}[b]
    \int_{x_0}^{x_n}\; y\,dx &= \int_{x_0}^{x_1} \; y\,dx + \int_{x_1}^{x_2} \; y\,dx + .... + \int_{x_{n-1}}^{x_n} \; y\,dx \\[1mm]
                             &= \frac{h}{2} \, (y_0 + y_1) + \frac{h}{2} \, (y_1 + y_2) + .... + \frac{h}{2} \, (y_{n-1} + y_n) \\[1mm]
                             &= \frac{h}{2} \, \{ y_0 + 2(y_1 + y_2 + ... + y_{n-1}) + y_n \}
  \end{aligned}
\end{equation}
\textbf{Question:} Illustrate the trapezoidal method geometrically.

@@

\subsection{Exercise}
\begin{enumerate}
\item For the given $(x,y)$ data points: Evaluate $\int_0^2 \, ydx$ using the trapezoidal rule.\\  $(0.0,0.399),\, (0.5,0.352),\,(1.0,0.242),\, (1.5,0.129),\,(2.0,0.054)$.

\item Evaluate using Simpson's $1/3$ rule:
  \begin{enumerate}
  \item $\displaystyle \int_1^3 \, \frac{1}{x}\,dx$  with $8$ stripes.
  \item $\displaystyle \int_0^{\pi /2} \, \sqrt{sinx}\, dx$ with $h= \pi /12$.

  \item $\displaystyle \int_3^7 \, x^2logx\, dx $ with $h= 1$.
    \item $\displaystyle \int_0^1 \, e^{-x^2} \, dx$ with $h=0.1$
  \end{enumerate}

\item Evaluate using Simpson's $3/8$ rule: $\displaystyle \int_0^1 \, \frac{dx}{1+x}$ with $h=1/6$.

\item Compute the values of $I = \displaystyle \int_0^1 \, \frac{dx}{1+x^2}$ using the trapezoidal rule with $h=0.5, \; 0.25, \; 0.125$. Then obtain a better estimate using \textbf{Romberg's method}.

\item Compute $\displaystyle I\left (\frac{\pi}{4} \right ), \; I\left (\frac{\pi}{8} \right ), \; I\left (\frac{\pi}{8}, \frac{\pi}{4} \right )$ for : $I = \displaystyle \int_0^{\pi/4} \, cos^2x\, dx$.

\item Using Trapezoidal rule and Simpson's $1/3$ rule evaluate the double integral $$\int_{_2}^2 \, \int_0^4 \; (x^2 -xy + y^2) \; dx\, dy.$$

\end{enumerate}
\subsection{Boole's and Weddle's Rules, \textit{Extra material}}
To incorporate the fourth order difference as well we need $5$ data points $x_0, ..., x_4$ and we need to integrate from $x_0$ to $x_4$. Then we obtain Boole's formula:
\begin{equation}
  \label{Boole's}
  \displaystyle \int_{x_0}^{x_4} \; y\,dx = \frac{2h}{45}\, (7y_0 + 32y_1 + 12y_2+32y_3+7y_4)
\end{equation}
If we incorporate differences u to six order then we obtain Weddle's formula as follows:
\begin{equation}
  \label{Weddle's}
  \displaystyle \int_{x_0}^{x_6} \; y\,dx = \frac{3h}{10}\, (y_0 + 5y_1 + y_2+6y_3+y_4+5y_5+y_6)
\end{equation}

\subsection{Romberg Integration}
This method improves the approximation obtained from the other methods. Here we describe the Romberge method to improve the results obtained from the Trapezoidal method. \\[2mm]
Suppose the integral $\displaystyle \int_a^b \, y\,dx$ is evaluated by the Trapezoidal rule with two different subintervals of widths $h_1$ and $h_2$, to obtain the approximate values $I_1$ and $I_2$, respectively. Romberg's method uses this two approximations: $I_1$ and $I_2$ to obtain the next better approximations which is given by
\begin{equation}
  \label{romberg}
I_3 = \frac{I_1h_2^2-I_2h_1^2}{h_2^2 - h_1^2}
\end{equation}

\subsection{Gaussian Quadrature}
The methods in the previous sections are based upon the values of function a equally-spaced points of the interval. Gauss derived a formula for which the points in the interval \textbf{need not be equally spaced}.
The Gaussian formula is expressed as:
\begin{equation}
  \label{gaussian}
\int_{-1}^{1} \, F(u)\, du = \sum_{i=1}^n \; W_i\, F(u_i)
\end{equation}
where $W_i$ and $u_i$ are called the \textit{weights} and \textit{abscissae}, respectively. It can be shown that the $u_i$ are the zeros of the $(n+1)^{th}$ \textbf{Legendre} polynomial $P_{n+1}(u)$ which can be generated usingthe recurrence relation:
\begin{equation*}
  (n+1)\,P_{n+1}(u) = (2n+1)\,u\,P_{n}(u) -n\,P_{n-1}(u)
\end{equation*}
where $P_0(u)=1$ and $P_1(u)=u$. It can also be shown that corresponding weights $W_i$ are given by
\begin{equation*}
  W_i =\int_{-1}^{1}\; \prod_{j=0, j\neq i}^n \; \frac{u-u_j}{u_i - u_j}\, du
\end{equation*}
However the weights and the abscissae are extensively tabulated for different values of $n$. We list the weights and the abscissae for values of $n$ up to $n=6$. \\[1mm]
\textbf{Change of Limits}
In the general case, the limits of the integral have to be changed to that of ~\ref{gaussian} by means of the following variable transformation:
\begin{equation*}
  x = \frac{1}{2} \, u(b-a) + \frac{1}{2}\, (a+b)
\end{equation*}

\begin{figure}[h]
  \centering
\includegraphics[width=12cm, height=9cm]{gaussian.png}
\end{figure}
\subsection{Exercise}
\begin{enumerate}
\item Derive the Gauss quadrature formula for $n=2$ and apply it to evaluate the integral $\displaystyle \int_{-1}^1 \, \frac{dx}{1+x^2}$.

\item Use the three point Gauss formula to evaluate the integral $\displaystyle \int_0^1 \, \frac{dx}{1+x}$.
\end{enumerate}
\section{Double Integration}
Formulae for the evaluation of double integral can be obtained by repeatedly applying the trapezoidal and Simpson's rules derived earlier.
\subsection{Trapezoidal Rule}
\begin{equation*}
  I = \int_{y_j}^{y_{j+1}}\; \int_{x_i}^{x_{i+1}} \; f(x,y)\;dx\,dy
\end{equation*}
where, $x_{i+1} = x_i + h$ and $y_{j+1} = y_j + k$.
By repeated application of trapezoidal rule to we get,
\begin{equation}
  \label{trapdouble}
  \begin{aligned}
    I &= \frac{h}{2} \,\int_{y_j}^{y_{j+1}}\; [f(x_i,y) + f(x_{i+1},y) ]\;dx\,dy \\[1mm]
      &= \frac{hk}{4} \, [f(x_i,y_j) + f(x_{i+1},y_j) + f(x_i,y_{j+1}) + f(x_{i+1},y_{j+1}) ]\;dx\,dy \\[1mm]
      &= \frac{hk}{4} \, [f_{i,j} + f_{i+1, j} + f_{i,j+1} + f_{i+1,j+1} ]\;dx\,dy
  \end{aligned}
\end{equation}

@@

\end{document}
